RationalGRL: A Framework for Argumentation and Goal Modeling

### 2019-02-04 Response to reviewers3.txt

We thank the reviewers for their work.

Reviewer 1 again proposes to accept the paper. Reviewer 2 recommends rejection. The editor proposes that we perform major revisions.

We have worked hard to improve the paper since the first submission. We propose a new framework, inserting argument schemes and formal argument semantics into goal modelling in a way that hasn't been done before. We formally ground this framework in a logical framework for argumentation that has been accepted as a de-facto standard for non-monotonic argumentation inside and outside the field of AI. We give a simple development process, and a fully-working tool. We have further performed an extensive qualitative emprirical coding study (which we call case study in our article), plus a modest expert survey.

One remark that has been made since the beginning is that, because we propose a new process/language/tool, we should test these in an industrial-strength case-study. However, we remain convinced that a large case study or experiment (where e.g. one compares test and control groups) would take an unreasonable amount of time and space given the material that is alreadty in the paper. See e.g. Estrada et al. (2006) "An empirical evaluation of the i* framework in a model-based software generation environment", who took 9 months to conduct the case study. Furthermore, there are many papers on goal-modeling and related subjects in REJ which perform less empirical evaluation than we do, and the few that do have industrial-strength case studies evaluate a modeling language or tool that was introduced in an earlier paper. We are not directly aware of papers on goal modeling languages doing extensive experimentation with test and control groups. Hence, we believe that asking us to do an industrial strength case-study or test-control group experiment is unreasonable for this paper, more so because this would likely yield enough new insights for a completely new paper.

@@

======= COMMENT #4
A theoretical problem of all the results reported here is the use of statistics like mean (here obviously called "average") and standard deviation based on data  on a Likert scale, since this is not permitted according to Measurement Theory. (I am aware, that this is often done, anyway, especially in practice, but the theoretical problem remains, especially for results in a prestigious scientific journal.) Such statistics could and should be done based on measured data, which are usually on an absolute scale.
In addition, has the probability distribution been tested before applying the t-test?
The specific results for Q6 are really strange.
Finally, as indicated above already, the results on usability of the tool are virtually useless.
======= REPLY #4
There is actually quite some discussion about whether Likert scales can be analysed as interval data - in academia more and more researchers (who publish in prestigious statistics journals) are starting to agree that one is allowed to do this. However, we agree that for our data it is perhaps better to err on the side of caution and use a non-parametric test which makes as little assumptions as possible (i.e. the sign test). Interestingly, the results are not much different: for Q2 (prev. Q6) we now cannot reject H0, where previously we could, but in the previous version (with the t-tests) we already indicated that Q2/Q6 has results which are clearly different from the other related questions.

======= COMMENT #5
The first argument on internal validity is strange and really not convincing. If the modeling task given in the experiment was not realistic (and this is most likely the case), then the case study does not help, see below.
I agree that a comparison with a control group is necessary!
-       Discussion
And yes,  a realistic study on this approach needs to involve asking CQs in the course of a development process.
Why did you not let users being guided by your process in the course of a study, when users like that? This would have provided evidence or at least insights on the usefulness of this process, while it is just a claim as it stands.
======= REPLY #5
Again: yes there are many ways in which we could have done our experiments and surveys differently, better, bigger etc. However, the reviewer makes it sound as if the current survey is next to useless, and we strongly disagree with this. It is a modest survey with a fairly high response rate, particularly as all 16 participants have significant expertise in software and requirements engineering - we could do an experiment with 80 inexperienced undergraduates, but what would that really tell us? The answers to the survey questions point in interesting future directions, and overall the experiment gives support to the idea that adding arguments to goal models is the way to go.

@@

======= COMMENT #6
In this context, let me also say a few words on the "Case Study":
-       It cannot be taken as evidence for the approach, since studying recorded design sessions (without goal modeling) is reported in this paper as the basis for creating the argumentation structures in the first place. And this was done by the authors themselves based on their intuition.
-       Actually, this is not really a case study in the sense used in scientific literature. As a matter of fact, it also does not contain lessons learned.
======= REPLY #6
We agree that "case study" is perhaps not the most opportune name. It is, if you want to be precise, a qualitative empirical coding study with a priori coding. We have changed the text in the paper to reflect this.

======= COMMENT #7
In addition, I am still not convinced by the argument for dismissing formal reasoning in propositional logic. After all, it is not theoretically difficult and tools exist. So, what is the point of presenting special-purpose "algorithms" here?
======= REPLY #7
The argumentation formalism we propose is a non-monotonic logic, that is, earlier results might not be true when new information is added. Formal reasoning in propositional logic, taking the semantics of classical logic, is inherently monotonic: you can add propositions (arguments, goals), but these will never change the status (true/false) of existing propositions. So what is needed is a new algorithm that shows what happens if new information is added, we cannot simply take an existing tool (theorem prover) for classical logic.

======= COMMENT #8
Finally, "defending" a binary framework simply by a reference is insufficient. Since the authors propose such a framework, they need to clearly point out its limitations in this regard.
======= REPLY #8
We already included some discussion in the previous revision. We have included an extra sentence on why we think numerical, more fine-grained frameworks are not very suitable. Furthermore, note that the framework is not "binary" in the sense that arguments can only be true or false (as in classical logic): arguments which are at one point considered defeated or overruled can later become justified again if new arguments are added.

======= COMMENT #9
From an editorial perspective, a careful proof reading of grammar would be necessary.
======= REPLY #9
This remark makes it sound as if there are *many* grammatical mistakes in the paper.

### 2019-02-04 Response to reviewers3.txt

We thank the reviewers for their work.

Reviewer 1 again proposes to accept the paper. Reviewer 2 recommends rejection. The editor proposes that we perform major revisions.

We have worked hard to improve the paper since the first submission. We propose a new framework, inserting argument schemes and formal argument semantics into goal modelling in a way that hasn't been done before. We formally ground this framework in a logical framework for argumentation that has been accepted as a de-facto standard for non-monotonic argumentation inside and outside the field of AI. We give an simply development process, and a fully-working tool. We have further performed an extensive coding analysis, plus a modest expert survey.

One remark that has been made since the beginning is that, because we propose a new process/language/tool, we should test these in an industrial-strength case-study. However, we remain convinced that a large case study or experiment (where e.g. one compares test and control groups) would take an unreasonable amount of time and space given the material that is already in the paper. See e.g. Estrada et al. (2006) "An empirical evaluation of the i* framework in a model-based software generation environment", who took 9 months to conduct the case study. Furthermore, there are many papers on goal-modeling and related subjects in REJ that perform less empirical evaluation than we do, and the few that do have industrial-strength case studies evaluate a modeling language or tool that was introduced earlier. We are not directly aware of papers on goal modeling languages doing extensive experimentation with test and control groups. Hence, we believe that asking us to do an industrial strength case-study or test-control group experiment is unreasonable for this paper, more so because this would likely yield enough new insights for a completely new paper.

The (smaller) remarks made by reviewer 2 since Revision 1 have been helpful, and we try to accommodate as many of them as possible. We would like to note that we have never simply refused suggestions or ignored comments. In some cases, we have replied at length to the reviewer in the response letter and have not added anything to the paper, but we do not think this counts as a refusal to take comments into account.

@@

Below we respond to reviewer 2's comments on the latest version of the article in detail.

======= COMMENT #1
First, there are "broad" and "general" hypotheses given, and based on them Null and Alternative Hypotheses. Still, the latter do not contain the level of significance for accepting or rejecting them based on statistical data and their analyses. (It has to be defined before performing an experiment.)
======= REPLY #1
We have added the significance levels below the hypothesis

======= COMMENT #2
H3 (and its specific counterparts) are strange in their formulation, since usability is not difficult, but use is difficult and usability potentially bad. In English, after "neither" there should be a "nor".
The paragraph apparently trying to motivate H3 sounds actually more like motivating user experience (UX) rather than usability, since testing the latter is first of all more about finding usability problems, something never mentioned in this paper. Also the related Null and Alternative Hypotheses are strange with regard to the given motivation. Much as in my previous review, I really wonder about including usability of such a prototype tool here. (As given later in the paper, this tool does not even have an Undo, and the results were completely inconclusive, anyway.)
======= REPLY #2
Our reasons for including the questions and hypothesis on usability (or UX) were not so that we could argue that said UX is particularly good (or bad). We wanted to show that the usability or UX of the tool did, in our opinion, not interfere greatly with what the participants filled in for the other questions. If a tool is very bad (keeps crashing, impossible to use, unclear interface), participants might be more inclined to be very negative about the rest (i.e. the RationalGRL features) as well. In any case, to avoid confusion we deleted H3 and only mention the user experience questions in passing.

======= COMMENT #3
For a comparison, actually a user study would have had to be designed, with independent and dependent variables, and participants trying both approaches in two groups (within-subject study design). In such a setting, something should be really measured in addition to subjective questionnaires. Much as indicated by the participants of the reported study (and as known from the literature), a key point of such argumentation systems is the overhead of providing them. There needs to be a return on invest, ideally in a balanced trade-off. I think that this should have been studied, since I am very much in favor of documenting rationale, but in practice developers will only do that if there is a clear benefit in return. The communication of opinions to others was obviously not really included.
======= REPLY #3
As we argue above, we think this is an unreasonable amount of work to ask given what is already in the paper.

@@

======= COMMENT #4
A theoretical problem of all the results reported here is the use of statistics like mean (here obviously called "average") and standard deviation based on data  on a Likert scale, since this is not permitted according to Measurement Theory. (I am aware, that this is often done, anyway, especially in practice, but the theoretical problem remains, especially for results in a prestigious scientific journal.) Such statistics could and should be done based on measured data, which are usually on an absolute scale.
In addition, has the probability distribution been tested before applying the t-test?
The specific results for Q6 are really strange.
Finally, as indicated above already, the results on usability of the tool are virtually useless.
======= REPLY #4
There is actually quite some discussion about whether Likert scales can be analysed as interval data - in academia more and more researchers (who publish in prestigious statistics journals) are starting to agree that one is allowed to do this. However, we agree that for our data it is perhaps better to err on the side of caution and use a non-parametric test which makes as little assumptions as possible (i.e. the sign test). Interestingly, the results are not much different: for Q2 (prev. Q6) we now cannot reject H0, where previously we could, but in the previous version (with the t-tests) we already indicated that Q2/Q6 has results which are clearly different from the other related questions.

======= COMMENT #5
The first argument on internal validity is strange and really not convincing. If the modeling task given in the experiment was not realistic (and this is most likely the case), then the case study does not help, see below.
I agree that a comparison with a control group is necessary!
-       Discussion
And yes,  a realistic study on this approach needs to involve asking CQs in the course of a development process.
Why did you not let users being guided by your process in the course of a study, when users like that? This would have provided evidence or at least insights on the usefulness of this process, while it is just a claim as it stands.
======= REPLY #5
We agree that are many ways in which we could have done our experiments and surveys differently and more extensively. However, we disagree with the insinuation that the current survey is next to useless. We provide a modest survey with a fairly high response rate, particularly as all 16 participants have significant expertise in software and requirements engineering. Alternatively, it would be possible to do an experiment with a large number of inexperienced undergraduate students, but we believe our current study provides us with more insights. Overall, the experiment gives modest support to the idea that adding arguments to goal models is useful and insightful for modelers.

@@

======= COMMENT #6
In this context, let me also say a few words on the "Case Study":
-       It cannot be taken as evidence for the approach, since studying recorded design sessions (without goal modeling) is reported in this paper as the basis for creating the argumentation structures in the first place. And this was done by the authors themselves based on their intuition.
-       Actually, this is not really a case study in the sense used in scientific literature. As a matter of fact, it also does not contain lessons learned.
======= REPLY #6
We agree that "case study" is perhaps not the most opportune name. It is, if you want to be precise, a qualitative empirical coding study with a priori coding. We have changed the text in the paper to reflect this.

======= COMMENT #7
In addition, I am still not convinced by the argument for dismissing formal reasoning in propositional logic. After all, it is not theoretically difficult and tools exist. So, what is the point of presenting special-purpose "algorithms" here?
======= REPLY #7
The argumentation formalism we propose is a non-monotonic logic, that is, earlier results might not be true when new information is added. Formal reasoning in propositional logic, taking the semantics of classical logic, is inherently monotonic: you can add propositions (arguments, goals), but these will never change the status (true/false) of existing propositions. Formal reasoning in propositional logic would be possible using, for instance, SAT solvers, but this would require a monotonic logic. Theorem provers for non-monotonic logics are significantly more complex, and not the focus of our research.

@@

======= COMMENT #8
Finally, "defending" a binary framework simply by a reference is insufficient. Since the authors propose such a framework, they need to clearly point out its limitations in this regard.
======= REPLY #8
We already included some discussion in the previous revision. We have included an extra sentence on why we think numerical, more fine-grained frameworks are not very suitable.

======= COMMENT #9
From an editorial perspective, a careful proof reading of grammar would be necessary.
======= REPLY #9
We have revised our entire article in detail and correct a large number of grammatical mistakes.

### 2019-02-04 abstract.tex
final changes

\begin{abstract}
Goal-oriented requirements modeling approaches aim at capturing intentions of various stakeholders involved in the development of an information system and refining them into low level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, by specifying alternative solutions for a goal, not all of the arguments can be found back in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. In this paper, we apply argumentation techniques from artificial intelligence to a goal modeling approach by using a case study in which stakeholders discuss requirements for a Traffic Simulator. We combine a simplified version of a traditional goal modeling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model. We also define the RationalGRL development process to create a RationalGRL model.

>>>>

\begin{abstract}
Goal-oriented requirements modeling approaches aim at capturing intentions of various stakeholders involved in the development of an information system and refining them into low level goals and tasks. The process of constructing a goal model usually involves discussions between a requirements engineer and a group of stakeholders. While it is possible to capture part of this discussion process in a goal model, by specifying alternative solutions for a goal, not all of the arguments can be found back in the resulting model. For instance, the discussion on whether to accept or reject a certain goal and the ultimate rationale for acceptance or rejection cannot be captured in current goal modeling languages. In this paper, we apply argumentation techniques from artificial intelligence to a goal modeling approach by using a coding analysis in which stakeholders discuss requirements for a Traffic Simulator. We combine a simplified version of a traditional goal modeling approach, the Goal-oriented Requirements Language (GRL), with a formal Practical Reasoning Argument Scheme (PRAS) for reasoning about goals into a new framework (RationalGRL). RationalGRL provides a formal semantics and tool support to capture the discussions and outcomes of the argumentation process that leads to a goal model. We also define the RationalGRL development process to create a RationalGRL model.

### 2019-02-04 background.tex
final changes

In this section, we first introduce our running example, after which we give a brief overview of the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}, which is the goal modeling language we use to integrate with the argumentation framework\footnote{For simplicity's sake, we consider a slightly cut-down version of GRL. All the core elements of GRL are included but, for example, we have included only one level of contribution strength, whereas GRL contains different levels of positive and negative contribution. This simplification does not in any way influence the case study. Furthermore, extending the GRL part of RationalGRL to include the full set of GRL elements is trivial.}. Next, we discuss the \emph{practical reasoning argument scheme (PRAS)}~\cite{atkinson2007}, an argument scheme that is used to form arguments and counter-arguments about situations involving goals, and we give informal examples of how argument and counterargument can influence the status of beliefs about goals.   %This will be our starting point in the next section.

@@

Our examples and case study are based on the data produced by a recent series of experiments by Tang et al. \cite{TangEtal2018}, who in turn base their work on the so-called Irvine experiment~\cite{UCIworkshop}. This experiment contains a well-known design reasoning assignment in software engineering about a traffic simulator. In this assignment (see Appendix~\ref{sect:designprompt}), designers are provided with a problem description, requirements, and a description of the desired outcomes. The client of the project is Professor E, who teaches civil engineering courses at an American university. In order for the professor to teach students various theories concerning traffic (such as queuing theory), traffic simulator software needs to be developed in which students can create visual maps of an area, regulate traffic, and so forth. Tang et al. asked designers (groups of students) to discuss the requirements of this traffic simulator. These discussions were recorded and transcribed. We used these transcripts for an extensive case study on the basis of which we develop our RationalGRL framework. Furthermore, we also use the traffic simulator case as a running example throughout this paper.

@@

\begin{figure}[t]
\centering
\includegraphics[]{img/fig_AF1.pdf}
\caption{PRAS aArguments and Attacks in the Traffic Simulation Example}
\label{fig:pras:example}
\end{figure}

@@

Looking at PRAS and its critical questions, one can see how it could be used to argue about goals and actions or, more specifically, about goal models. In fact, in some of our previous work on RationalGRL \cite{vanzee-etal:renext2015,vanZee-etal:er2016} we used PRAS-arguments such as the ones above to capture reasoning about goals, and provided a translation to goal models. However, one problem of this approach is that we cannot literally use PRAS and its critical questions, as there are elements in the GRL language, such as actors and resources, which cannot be found in PRAS. Furthermore, it is not directly clear whether the critical questions as proposed by Atkinson and Bench-Capon~\cite{atkinson2007} actually apply to GRL models. In fact, our case study (Section \ref{sect:gmas}) shows that when discussing requirements, people very often do not structure their reasoning nicely in the way that PRAS presents it. That is, it is rarely the case that the discussants setting up an argument ``We have goal $G$, $A$ realizes $G$ \emph{Therefore} we should perform $A$''. A typical discussion is much more unstructured, as is clear from the transcript excerpts in Appendix~\ref{sect:transcripts:excerpts}. Thus, if we would use the version of PRAS presented in this section for our argumentation, we would violate requirement 1: The argumentation techniques must capture the actual discussions of the stakeholders or designers in the early requirements engineering phase. Our solution is to develop our own set of argument schemes and critical questions by analyzing transcripts of discussions about the traffic simulator. This set of schemes and questions and our case study are described in the next section.

>>>>

In this section, we first introduce our running example, after which we give a brief overview of the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}, which is the goal modeling language we use to integrate with the argumentation framework.\footnote{For simplicity's sake, we consider a slightly cut-down version of GRL. All the core elements of GRL are included but, for example, we have included only one level of contribution strength, whereas GRL contains different levels of positive and negative contribution. This simplification does not in any way influence the coding analysis. Furthermore, extending the GRL part of RationalGRL to include the full set of GRL elements is straightforward.} Next, we discuss the \emph{practical reasoning argument scheme (PRAS)}~\cite{atkinson2007}, an argument scheme that is used to form arguments and counter-arguments about situations involving goals, and we give informal examples of how argument and counterargument can influence the status of beliefs about goals.   %This will be our starting point in the next section.

@@

Our examples and coding analysis are based on the data produced by a recent series of experiments by Tang et al. \cite{TangEtal2018}, who in turn base their work on the so-called Irvine experiment~\cite{UCIworkshop}. This experiment contains a well-known design reasoning assignment in software engineering about a traffic simulator. In this assignment (see Appendix~\ref{sect:designprompt}), designers are provided with a problem description, requirements, and a description of the desired outcomes. The client of the project is Professor E, who teaches civil engineering courses at an American university. In order for the professor to teach students various theories concerning traffic (such as queuing theory), traffic simulator software needs to be developed in which students can create visual maps of an area, regulate traffic, and so forth. Tang et al. asked designers (groups of students) to discuss the requirements of this traffic simulator. These discussions were recorded and transcribed. We performed a coding analysis on these transcripts, which we used to develop our RationalGRL framework. Furthermore, we also use the traffic simulator case as a running example throughout this paper.

@@

\begin{figure}[t]
\centering
\includegraphics[]{img/fig_AF1.pdf}
\caption{PRAS Arguments and Attacks in the Traffic Simulation Example}
\label{fig:pras:example}
\end{figure}

@@

Looking at PRAS and its critical questions, one can see how it could be used to argue about goals and actions or, more specifically, about goal models. In fact, in some of our previous work on RationalGRL \cite{vanzee-etal:renext2015,vanZee-etal:er2016} we used PRAS-arguments such as the ones above to capture reasoning about goals, and provided a translation to goal models. However, one problem of this approach is that we cannot literally use PRAS and its critical questions, as there are elements in the GRL language, such as actors and resources, which cannot be found in PRAS. Furthermore, it is not directly clear whether the critical questions as proposed by Atkinson and Bench-Capon~\cite{atkinson2007} actually apply to GRL models. In fact, our coding analysis (Section \ref{sect:gmas}) shows that when discussing requirements, people very often do not structure their reasoning nicely in the way that PRAS presents it. That is, it is rarely the case that the discussants setting up an argument ``We have goal $G$, $A$ realizes $G$ \emph{Therefore} we should perform $A$''. A typical discussion is much more unstructured, as is clear from the transcript excerpts in Appendix~\ref{sect:transcripts:excerpts}. Thus, if we would use the version of PRAS presented in this section for our argumentation, we would violate Requirement 1: The argumentation techniques must capture the actual discussions of the stakeholders or designers in the early requirements engineering phase. Our solution is to develop our own set of argument schemes and critical questions by analyzing transcripts of discussions about the traffic simulator. This set of schemes and questions and our coding analysis are described in the next section.

### 2019-02-04 casestudy.tex
final changes

\section{Argument Schemes for Goal Modeling: a Case Study}
\label{sect:gmas}

Recall that \textbf{requirement 1} of our RationalGRL framework is that the argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase. To get a sense of such discussions, we performed a case study to examine which types of discourse are used during discussions of system requirements, and how these discourse types can be captured as argument schemes and critical questions. We manually coded transcripts of such discussions using a list of argument schemes and critical questions based on GRL and PRAS. In this section, we present our case study. All original transcripts, coding, and models are available in our online repository, which can be found at:

@@

In order to obtain actual requirements discussions, we turned to a recent series of experiments by Tang et al. \cite{TangEtal2018}. In these experiments, 12 groups of two or three students in a Software Architecture course at MSc level were given the traffic simulator assignment (Appendix~\ref{sect:designprompt}). These groups had a maximum of two hours to design a traffic simulator, which included a discussion of the requirements of this traffic simulator. The students did not use any goal modeling technique in the course or during the discussions. They were asked to record their design sessions, and the recordings were subsequently transcribed. We used three of these transcripts, totaling 153 pages, for our case study.

>>>>

\section{Argument Schemes for Goal Modeling: a Coding Analysis}
\label{sect:gmas}

Recall that \textbf{Requirement 1} of our RationalGRL framework is that the argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase. To get a sense of such discussions, we performed a coding analysis to examine which types of discourse are used during discussions of system requirements, and how these discourse types can be captured as argument schemes and critical questions. We manually coded transcripts of such discussions using a list of argument schemes and critical questions based on GRL and PRAS. In this section, we present our coding analysis. All original transcripts, codings, and models are available in our online repository, which can be found at:

@@

In order to obtain actual requirements discussions, we turned to a recent series of experiments by Tang et al. \cite{TangEtal2018}. In these experiments, 12 groups of two or three students in a Software Architecture course at MSc level were given the traffic simulator assignment (Appendix~\ref{sect:designprompt}). These groups had a maximum of two hours to design a traffic simulator, which included a discussion of the requirements of this traffic simulator. The students did not use any goal modeling technique in the course or during the discussions. They were asked to record their design sessions, and the recordings were subsequently transcribed. We used three of these transcripts, totaling 153 pages, for our coding analysis.

### 2019-02-04 discussion.tex
final changes

\paragraph{Empirical validation}
We have performed a case study and found more than 200 incidences of arguments and questions. We also performed an experiment with 16 participants, testing the usability and usefulness of RationalGRL compared to other goal modeling languages. However, what is still lacking is an industrial-strength case study testing extensively whether, and how, the tools and techniques provided by the RationalGRL framework really improve early phase requirements engineering. The complexity of the domain makes focused experiments difficult, but not impossible. For example, both Murakannaiah \emph{et al.}~\cite{murukannaiah2015} and Tang \emph{et al.}~\cite{TangEtal2018} provided a realistic but bounded problem for their subjects to solve, and similar experiments are imaginable to test RationalGRL: have two groups solve a goal modeling problem, one using only GRL and one using RationalGRL, and examine how they perform. Naturally, the type of problem will greatly influence the outcome. For a simple problem which two people have to solve in an hour, RationalGRL will most likely be of little benefit. However, a more complex problem on which multiple persons have to work asynchronously might benefit from the extra tools offered by RationalGRL. Furthermore, the outcome also depends on which part of the set of tools and techniques offered by RationalGRL is tested: the development process, the tool, or a list of critical questions to serve as reminders during the goal modeling process.

@@

Our list of argument schemes and critical questions in Table~\ref{table:argument-schemes} was constructed by performing an extensive case study in which we analyzed 153 pages of transcripts of discussions among designers of a traffic simulator information system. In this way, we ensure that the argumentation techniques capture the actual discussions of the stakeholders or designers in the early requirements engineering phase (\textbf{requirement 1}).

>>>>

\paragraph{Empirical validation}
We have performed a coding analysis and found more than 200 occurrences of arguments and questions. We also performed an experiment with 16 participants, testing the usability and usefulness of RationalGRL compared to other goal modeling languages. However, what is still lacking is an industrial-strength case study testing extensively whether, and how, the tools and techniques provided by the RationalGRL framework really improve early phase requirements engineering. The complexity of the domain makes focused experiments difficult, but not impossible. For example, both Murakannaiah \emph{et al.}~\cite{murukannaiah2015} and Tang \emph{et al.}~\cite{TangEtal2018} provided a realistic but bounded problem for their subjects to solve, and similar experiments are imaginable to test RationalGRL: have two groups solve a goal modeling problem, one using only GRL and one using RationalGRL, and examine how they perform. Naturally, the type of problem will greatly influence the outcome. For a simple problem which two people have to solve in an hour, RationalGRL will most likely be of little benefit. However, a more complex problem on which multiple persons have to work asynchronously might benefit from the extra tools offered by RationalGRL. Furthermore, the outcome also depends on which part of the set of tools and techniques offered by RationalGRL is tested: the development process, the tool, or a list of critical questions to serve as reminders during the goal modeling process.

@@

Our list of argument schemes and critical questions in Table~\ref{table:argument-schemes} was constructed by performing a coding analysis in which we analyzed 153 pages of transcripts of discussions among designers of a traffic simulator information system. In this way, we ensure that the argumentation techniques capture the actual discussions of the stakeholders or designers in the early requirements engineering phase (\textbf{requirement 1}).

### 2019-02-04 formalframework.tex
final changes

In Section~\ref{sect:overview}, we have shown through a language definition and informal examples from our case study that it is possible to trace elements of a goal model back to their underlying arguments (\textbf{requirement 2}), and that it is possible to determine the effect of changes in the underlying argumentation on a goal model, and vice versa (\textbf{requirement 3}). A formal rendition of this traceability will be presented in this section, in which we present a logical formalization of the RationalGRL framework. This is done for multiple reasons: (i) Most approaches in formal argumentation (cf. Section~\ref{sect:background:pras}) use formal logic, allowing us to employ existing technique directly in order to compute which arguments are accepted and which are rejected, (ii) we can be more precise about how critical questions are answered, (iii) we can show that RationalGRL models can be translated to valid GRL models and vice versa in a precise way, and (iv) the formal approach is a basis for automating the framework in terms of tool support, which we present in the next section.

>>>>

In Section~\ref{sect:overview}, we have shown through a language definition and informal examples from our coding analysis that it is possible to trace elements of a goal model back to their underlying arguments (\textbf{Requirement 2}), and that it is possible to determine the effect of changes in the underlying argumentation on a goal model, and vice versa (\textbf{Requirement 3}). A formal rendition of this traceability will be presented in this section, in which we present a logical formalization of the RationalGRL framework. This is done for multiple reasons: (i) Most approaches in formal argumentation (cf. Section~\ref{sect:background:pras}) use formal logic, allowing us to employ existing technique directly in order to compute which arguments are accepted and which are rejected, (ii) we can be more precise about how critical questions are answered, (iii) we can show that RationalGRL models can be translated to valid GRL models and vice versa in a precise way, and (iv) the formal approach is a basis for automating the framework in terms of tool support, which we present in the next section.
